---
title: "BikeShare Analysis"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## 1. Load packages

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(ggplot2)
```

2.  Read in data and merge tables

```{r, 'results=hide'}
setwd("/Users/alishalaby/MyDocuments/Project Portfolio/BikeShareAnalysis/Data")

bike_data <- rbind(
  read_csv("202009-divvy-tripdata.csv"), 
  read_csv("202010-divvy-tripdata.csv"),
  read_csv("202011-divvy-tripdata.csv"),
  read_csv("202012-divvy-tripdata.csv"),
  read_csv("202101-divvy-tripdata.csv"),
  read_csv("202102-divvy-tripdata.csv"),
  read_csv("202103-divvy-tripdata.csv"),
  read_csv("202104-divvy-tripdata.csv"),
  read_csv("202105-divvy-tripdata.csv"),
  read_csv("202106-divvy-tripdata.csv"),
  read_csv("202107-divvy-tripdata.csv"),
  read_csv("202108-divvy-tripdata.csv"))
```

## 2. Process Data

3.  Understand data structure

```{r}
head(bike_data)
str(bike_data)
```

4.  Check NA values

```{r}
sum(is.na(bike_data)) #total number of NA values
bike_data %>% summarise_all(~ sum(is.na(.))) #number of NA values per column
```

Evident that most na values are contained in station-based: 1. start_station_name 2. start_station_id 3. end_station_name 4. end_station_id We should leave the data as is and not replace NA values

5.  Drop 'lat' and 'long' columns

```{r}
bike_data <- bike_data %>% 
  select(-c(start_lat, start_lng, end_lat, end_lng))
```

6.  Add columns 'weekday', 'ride_length' and 'month'

```{r}
bike_data <- 
  bike_data %>% 
  mutate(weekday = weekdays(as.Date(bike_data$started_at))) %>%
  mutate(ride_length = ended_at - started_at) %>% 
  mutate(month = months(as.Date(bike_data$started_at)))
```

7.  Convert 'ride_length' to minutes from second

```{r}
glimpse(bike_data$ride_length)
bike_data$ride_length <- as.numeric(bike_data$ride_length)
bike_data$ride_length <- as.numeric(bike_data$ride_length/60)
```

8.  Filter 'bad' data

```{r}
bike_data <- bike_data %>%
  filter(ride_length >1) %>% #bikes with more than 1 min of use
  filter(ride_length <= 1440) # bikes atleast 1 day of use
```

## 3. Analyze

9.  Summary statistics

```{r}
summary(bike_data$ride_length) 
```

10. Aggregate data based on casual vs member riders

```{r}
bike_data %>% 
  group_by(member_casual) %>% 
  summarise(number_of_riders = n(), 
            mean = mean(ride_length),
            median = median(ride_length))
```

11. Average ride time day of the week members vs casuals

```{r}
rider_weekday <- bike_data %>% 
  group_by(member_casual, weekday) %>% 
  summarise(average_ride_length = mean(ride_length), number_of_rides = n())
  
rider_weekday$weekday <- factor(rider_weekday$weekday, levels= c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
rider_weekday <- rider_weekday[order(rider_weekday$weekday),]
```

12. Analyze ridership data by type of vehicle

```{r}
bike_type <- bike_data %>%
  group_by(member_casual, rideable_type) %>%
  summarise(number_of_riders = n())
```

13. Analyze station data

```{r}
station_data <- bike_data %>% 
  group_by(member_casual, start_station_name) %>%
  summarise(number_of_stations = n()) %>% 
  arrange(desc(number_of_stations))

station_data <- station_data %>%
  rename(station_name = start_station_name)
  
```

14. Analyze ride_length by month

```{r}
rider_month <- bike_data %>% 
  group_by(member_casual, month) %>%
  summarise(number_of_rides = n(), average_ride_length = mean(ride_length))

rider_month$month <- factor(rider_month$month, levels= c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
rider_month <- rider_month[order(rider_month$month),]

```

## 4. Visualize

15. Visualize ridership distribution

```{r}
bike_data %>%
  group_by(member_casual) %>%
  summarize(number_of_rides = n()) %>%
  ggplot(aes(x="", y=number_of_rides, fill=member_casual)) +
  geom_col() + labs(title="Number of rides", x="", y="") +
  geom_text(aes(label = number_of_rides), position = position_stack(vjust = 0.5)) +
  coord_polar("y") 
```

15. Visualize weekly ridership

```{r}
rider_weekday %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) +
  geom_col(position="dodge") + 
  labs(title="Weekly Ridership", x="Week", y="Number of Rides", fill="User Type") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

16. Visualize weekly average ride length

```{r}
rider_weekday %>% 
  ggplot(aes(x=weekday, y=average_ride_length, fill=member_casual)) + 
  geom_col() +
  labs(title="Average Weekly Ride Length", x="Week", y="Average Ride Length", fill="User Type") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

17. Visualize monthly ridership

```{r}
rider_month %>%
  ggplot(aes(x=month, y=number_of_rides, fill=member_casual)) +
  geom_col(position="dodge") + 
  labs(title="Monthly Ridership", x="Month", y="Number of Rides", fill="User Type") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## 5. Export

18. 

```{r}
# Get total riders and bike types in order to visualize on Tableau
total_riders <- data.frame(table(bike_data$member_casual))
total_types <- data.frame(table(bike_data$rideable_type))

write_csv(total_riders, "total_riders.csv")
write_csv(total_types, "total_types.csv")
write_csv(station_data, "station_data.csv")
write_csv(bike_data, "bike_data.csv")
write_csv(bike_type, "bike_type.csv")
write_csv(rider_month, "rider_month.csv")
write_csv(rider_weekday, "rider_weekday.csv")

```
